Why shouldn't you use numerical tests to diagnose the linear regression models?

### 2022-05-05 paper_comparison/paper_comparison/paper_comparison.Rmd
clean up scripts
update scripts, processed data and result

<!-- week 3 -->
comparison of conventional tests:
1. power (visual test vs. conventional test)
(visual test most different one (everything test, any departure))
2. investigate the difference (gap), give examples
3. conventional is too sensitive
4. make conventional less sensitive (vary alpha)

@@

However, unlike confirmatory data analysis built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding an over-interpretation of the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015].

Recently, a new branch of statical inference

Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots [@gelman_exploratory_2004]. Later, in the Bayesian context, data plots was systematically considered as model diagnostics by taking advantage of the data simulated from the assumed statistical models [@gelman_bayesian_2003; @gelman_exploratory_2004].

>>>>

<!-- week 3 -->
comparison of conventional tests:
1. power (visual test vs. conventional test)
(visual test most different one (everything test, any departure))
plot figure in a paper, desc, exp
2. investigate the difference (gap), give examples
3. conventional is too sensitive
4. make conventional less sensitive (vary alpha)

@@

<!-- However, unlike confirmatory data analysis built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding an over-interpretation of the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015]. -->

<!-- Recently, a new branch of statical inference  -->

<!-- Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots [@gelman_exploratory_2004]. Later, in the Bayesian context, data plots was systematically considered as model diagnostics by taking advantage of the data simulated from the assumed statistical models [@gelman_bayesian_2003; @gelman_exploratory_2004].  -->

# Experimental design

# Data processing

# Results

## Overview of the Data

We collected 400 lineup evaluations made by 20 participants in experiment I and 880 lineup evaluations made by 44 participants in experiment II. In total, 442 unique lineups were evaluated by 64 subjects. In experiment I, one of the participants skipped all 20 lineups. Hence, the submission was rejected and removed from the dataset. In experiment II, there was a participant failed one of the two attention checks, but there was no further evidence of low-effort throughout the experiment. Therefore, the submission was kept.

## Power comparision

1. power (visual test vs. conventional test)
(visual test most different one (everything test, any departure))
plot figure in a paper, desc, exp
2. investigate the difference (gap), give examples
3. conventional is too sensitive
4. make conventional less sensitive (vary alpha)

It was expected that with larger effect size, both conventional test and visual test will have higher probability in rejecting the null hypothesis when it is not true. To explore the difference in power between conventional test and visual test, ten logistic regression models are fit to the data, with natural logarithm of the effect size as the only fixed effect, and whether the test successfully rejects the null hypothesis as the response variable. Table \ref{tab:powerglmheter} and \ref{tab:powerglmheter} summarizes the results. The natural logarithm of the effect size is positive and significant across all models. Figure \ref{fig:power-com} illustrates the fitted models, while providing the local constant estimate of the power of F-test and Breusch–Pagan test for comparison. Data for the conventional test is simulated under the same model setting used in two experiments and 500000 samples are drawn for both cubic and heteroskedasticity model.

From Figure \ref{fig:power-com}, it can be observed that the fitted power of visual test increased as the number of evaluations increased for both cubic and heteroskedasticity model.

For heteroskedasticity model, this phenomenon was more obvious as the curves of visual tests with evaluations greater than two were always above the curves of visual test with evaluations smaller than two. However, this only held for large enough effect size. For small effect size, visual tests with fewer evaluations might have greater power. Note that, the expected power of visual test derived by @majumder_validation_2013 followed by the assumption that human has the ability to select the plot with the highest t-statistic from a lineup under the classical linear model regression setting showed similar properties, where visual tests with fewer evaluations were expected to perform better when the parameter values close to the null hypothesis.

For cubic model, the separation between curves was small. Fitted power of visual test with three to five evaluations were almost identical to each other in regards of effect size. In addition, all five curves peaked at one as effect size increased, suggesting that identification of non-linearity as a visual task can be completed reliably by human when the effect size is large enough.

As shown in Figure \@ref(fig:power-com), both F-test and Breusch–Pagan test generally possessed greater power than visual test. It was also found that there was a huge gap between the conventional test curves and the visual test curves at around $log(\text{effect size}) = 0$ for the cubic model and $log(\text{effect size}) = 2.5$ for the heteroskedasticity model.

We further analysed the lineup with the corresponding effect size.

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)
```

```{r}
library(tidyverse)
library(visage)
pilot <- read_csv(here::here("data/processed/processed_pilot.csv"))
survey <- read_csv(here::here("data/processed/processed_surveys.csv"))
sim_conv <- read_csv(here::here("data/processed/sim_conv_result.csv"), col_types = "dddlc")
```

```{r}
survey <- survey %>%
  mutate(log_effect_size = log(effect_size))

sim_conv <- sim_conv %>%
  mutate(log_effect_size = log(effect_size))

lineup_info <- select(survey, type:answer, conventional_p_value, ori_difficulty, exp, exp_lineup_id, log_effect_size) %>%
  group_by(exp_lineup_id) %>% 
  filter(row_number() == 1) %>%
  ungroup()

write_csv(lineup_info, here::here("data/processed/processed_lineup_info.csv"))

# Local constant estimator
kernel_regression_cubic <- function(h = 0.3) {
  with(filter(sim_conv, type == "cubic", log_effect_size > -5, log_effect_size < 10),
     KernSmooth::locpoly(x = log_effect_size, y = reject, bandwidth = h, degree = 0, gridsize = 1000)) %>%
    as_tibble() %>%
    mutate(type = "cubic", h = h) %>%
    rename(log_effect_size = x) %>%
    rename(reject = y)
}

# Local constant estimator
kernel_regression_heter <- function(h = 0.4) {
  with(filter(sim_conv, type != "cubic", log_effect_size > -3),
     KernSmooth::locpoly(x = log_effect_size, y = reject, bandwidth = h, degree = 0, gridsize = 1000)) %>%
    as_tibble() %>%
    mutate(type = "heteroskedasticity", h = h) %>%
    rename(log_effect_size = x) %>%
    rename(reject = y)
}
```

```{r}
if (file.exists(here::here("data/processed/lineup_env.rds"))) {
  env <- readRDS(here::here("data/processed/lineup_env.rds"))
} else {
  env <- new.env()
}

save_lineup_env <- function(env) {
  saveRDS(env, here::here("data/processed/lineup_env.rds"))
}

survey_pvalue <- survey %>%
  calc_p_value_multi(lineup_id = "exp_lineup_id", 
                     detected = "detect", 
                     n_sel = "num_selection",
                     comb = TRUE,
                     n_eval = 1:5,
                     n_sim = 100000,
                     cache_env = env) %>%
  rename(exp_lineup_id = lineup_id) %>%
  unnest_longer(col = p_value) %>%
  left_join(lineup_info) %>%
  mutate(reject = eval_p_value(p_value, tol = 0.01))

save_lineup_env(env)
```

```{r results='asis'}
map(1:5, function(x) glm(reject ~ log_effect_size,
                         family = binomial(),
                         data = filter(survey_pvalue, n_eval == {{x}}, type == "cubic"))) %>%
  stargazer::stargazer(header = FALSE,
            label = "tab:powerglmcubic",
            title = "Logistic regressions with rejection as response variable and natural logarithm of the effect size as regressor fitted on the data collected from experiment I and II for cubic model. Five regression were fitted for different number of evaluations.",
            style = "default",
            column.labels = c("num\\_eval: 1", "num\\_eval:2", "num\\_eval: 3", "num\\_eval: 4", "num\\_eval:5"),
            model.numbers = FALSE)
```

```{r results="asis"}
map(1:5, function(x) glm(reject ~ log_effect_size,
                         family = binomial(),
                         data = filter(survey_pvalue, n_eval == {{x}}, type == "heteroskedasticity"))) %>%
  stargazer::stargazer(header = FALSE,
            label = "tab:powerglmheter",
            title = "Logistic regressions with rejection as response variable and natural logarithm of the effect size as regressor fitted on the data collected from experiment I and II for heteroskedasticity model. Five regression were fitted for different number of evaluations.",
            style = "default",
            column.labels = c("num\\_eval: 1", "num\\_eval:2", "num\\_eval: 3", "num\\_eval: 4", "num\\_eval:5"),
            model.numbers = FALSE)
```

```{r power-com, fig.cap="\\label{fig:power-com}power com"}
survey_pvalue %>%
  mutate(reject = as.numeric(reject)) %>%
  mutate(n_eval = factor(n_eval)) %>%
  ggplot() +
  geom_point(aes(log_effect_size, reject), alpha = 0.01) +
  geom_line(data = kernel_regression_cubic(), aes(log_effect_size, reject), linetype = 2) +
  geom_line(data = kernel_regression_heter(), aes(log_effect_size, reject), linetype = 2) +
  geom_smooth(aes(log_effect_size, reject, group = n_eval, col = n_eval),
              method = "glm", method.args = list(family = binomial), size = 0.5, se = FALSE) +
  scale_colour_viridis_d(option = "B", begin = 0.5, end = 0.9) +
  facet_wrap(~type, scales = "free_x") +
  ylab("Estimated power")
```

### 2022-05-05 paper_comparison/paper_comparison/paper_comparison.tex
update scripts, processed data and result

comparison of conventional tests: 1. power (visual test vs.~conventional
test) (visual test most different one (everything test, any departure))
2. investigate the difference (gap), give examples 3. conventional is
too sensitive 4. make conventional less sensitive (vary alpha)

@@

However, unlike confirmatory data analysis built upon rigorous
statistical procedures, e.g., hypothesis testing, visual diagnostics
relies on graphical perception - human's ability to interpret and decode
the information embedded in the graph \citep{cleveland_graphical_1984},
which is to some extent subjective. Further, visual discovery suffers
from its unsecured and unconfirmed nature where the degree of the
presence of the visual features typically can not be measured
quantitatively and objectively, which may lead to over or
under-interpretations of the data. One such example is finding an
over-interpretation of the separation between gene groups in a
two-dimensional projection from a linear discriminant analysis when in
fact there are no differences in the expression levels between the gene
groups and separation is not an uncommon occurrence
\citep{roy_chowdhury_using_2015}.

Recently, a new branch of statical inference

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots
\citep{gelman_exploratory_2004}. Later, in the Bayesian context, data
plots was systematically considered as model diagnostics by taking
advantage of the data simulated from the assumed statistical models
\citep{gelman_bayesian_2003, gelman_exploratory_2004}.

>>>>

comparison of conventional tests: 1. power (visual test vs.~conventional
test) (visual test most different one (everything test, any departure))
plot figure in a paper, desc, exp 2. investigate the difference (gap),
give examples 3. conventional is too sensitive 4. make conventional less
sensitive (vary alpha)

@@

\hypertarget{experimental-design}{%
\section{Experimental design}\label{experimental-design}}

\hypertarget{data-processing}{%
\section{Data processing}\label{data-processing}}

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{overview-of-the-data}{%
\subsection{Overview of the Data}\label{overview-of-the-data}}

We collected 400 lineup evaluations made by 20 participants in
experiment I and 880 lineup evaluations made by 44 participants in
experiment II. In total, 442 unique lineups were evaluated by 64
subjects. In experiment I, one of the participants skipped all 20
lineups. Hence, the submission was rejected and removed from the
dataset. In experiment II, there was a participant failed one of the two
attention checks, but there was no further evidence of low-effort
throughout the experiment. Therefore, the submission was kept.

\hypertarget{power-comparision}{%
\subsection{Power comparision}\label{power-comparision}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  power (visual test vs.~conventional test) (visual test most different
  one (everything test, any departure)) plot figure in a paper, desc,
  exp
\item
  investigate the difference (gap), give examples
\item
  conventional is too sensitive
\item
  make conventional less sensitive (vary alpha)
\end{enumerate}

It was expected that with larger effect size, both conventional test and
visual test will have higher probability in rejecting the null
hypothesis when it is not true. To explore the difference in power
between conventional test and visual test, ten logistic regression
models are fit to the data, with natural logarithm of the effect size as
the only fixed effect, and whether the test successfully rejects the
null hypothesis as the response variable. Table \ref{tab:powerglmheter}
and \ref{tab:powerglmheter} summarizes the results. The natural
logarithm of the effect size is positive and significant across all
models. Figure \ref{fig:power-com} illustrates the fitted models, while
providing the local constant estimate of the power of F-test and
Breusch--Pagan test for comparison. Data for the conventional test is
simulated under the same model setting used in two experiments and
500000 samples are drawn for both cubic and heteroskedasticity model.

From Figure \ref{fig:power-com}, it can be observed that the fitted
power of visual test increased as the number of evaluations increased
for both cubic and heteroskedasticity model.

For heteroskedasticity model, this phenomenon was more obvious as the
curves of visual tests with evaluations greater than two were always
above the curves of visual test with evaluations smaller than two.
However, this only held for large enough effect size. For small effect
size, visual tests with fewer evaluations might have greater power. Note
that, the expected power of visual test derived by
\citet{majumder_validation_2013} followed by the assumption that human
has the ability to select the plot with the highest t-statistic from a
lineup under the classical linear model regression setting showed
similar properties, where visual tests with fewer evaluations were
expected to perform better when the parameter values close to the null
hypothesis.

For cubic model, the separation between curves was small. Fitted power
of visual test with three to five evaluations were almost identical to
each other in regards of effect size. In addition, all five curves
peaked at one as effect size increased, suggesting that identification
of non-linearity as a visual task can be completed reliably by human
when the effect size is large enough.

As shown in Figure @ref(fig:power-com), both F-test and Breusch--Pagan
test generally possessed greater power than visual test. It was also
found that there was a huge gap between the conventional test curves and
the visual test curves at around \(log(\text{effect size}) = 0\) for the
cubic model and \(log(\text{effect size}) = 2.5\) for the
heteroskedasticity model.

We further analysed the lineup with the corresponding effect size.

\begin{table}[!htbp] \centering 
  \caption{Logistic regressions with rejection as response variable and natural logarithm of the effect size as regressor fitted on the data collected from experiment I and II for cubic model. Five regression were fitted for different number of evaluations.} 
  \label{tab:powerglmcubic} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
\cline{2-6} 
\\[-1.8ex] & \multicolumn{5}{c}{reject} \\ 
 & num\_eval: 1 & num\_eval:2 & num\_eval: 3 & num\_eval: 4 & num\_eval:5 \\ 
\hline \\[-1.8ex] 
 log\_effect\_size & 0.929$^{***}$ & 1.258$^{***}$ & 1.096$^{***}$ & 1.113$^{***}$ & 1.244$^{***}$ \\ 
  & (0.073) & (0.076) & (0.054) & (0.056) & (0.077) \\ 
  & & & & & \\ 
 Constant & $-$1.969$^{***}$ & $-$2.534$^{***}$ & $-$1.549$^{***}$ & $-$1.334$^{***}$ & $-$1.404$^{***}$ \\ 
  & (0.169) & (0.173) & (0.113) & (0.122) & (0.185) \\ 
  & & & & & \\ 
\hline \\[-1.8ex] 
Observations & 630 & 1,135 & 1,660 & 1,904 & 1,879 \\ 
Log Likelihood & $-$220.882 & $-$314.394 & $-$471.152 & $-$396.979 & $-$206.296 \\ 
Akaike Inf. Crit. & 445.764 & 632.788 & 946.305 & 797.958 & 416.592 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

\begin{table}[!htbp] \centering 
  \caption{Logistic regressions with rejection as response variable and natural logarithm of the effect size as regressor fitted on the data collected from experiment I and II for heteroskedasticity model. Five regression were fitted for different number of evaluations.} 
  \label{tab:powerglmheter} 
\begin{tabular}{@{\extracolsep{5pt}}lccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{5}{c}{\textit{Dependent variable:}} \\ 
\cline{2-6} 
\\[-1.8ex] & \multicolumn{5}{c}{reject} \\ 
 & num\_eval: 1 & num\_eval:2 & num\_eval: 3 & num\_eval: 4 & num\_eval:5 \\ 
\hline \\[-1.8ex] 
 log\_effect\_size & 0.821$^{***}$ & 1.104$^{***}$ & 1.157$^{***}$ & 1.387$^{***}$ & 1.651$^{***}$ \\ 
  & (0.086) & (0.079) & (0.066) & (0.076) & (0.110) \\ 
  & & & & & \\ 
 Constant & $-$4.472$^{***}$ & $-$5.887$^{***}$ & $-$5.431$^{***}$ & $-$6.154$^{***}$ & $-$7.123$^{***}$ \\ 
  & (0.434) & (0.415) & (0.344) & (0.400) & (0.591) \\ 
  & & & & & \\ 
\hline \\[-1.8ex] 
Observations & 630 & 1,138 & 1,596 & 1,662 & 1,416 \\ 
Log Likelihood & $-$333.119 & $-$631.519 & $-$835.317 & $-$650.570 & $-$342.058 \\ 
Akaike Inf. Crit. & 670.238 & 1,267.038 & 1,674.634 & 1,305.140 & 688.116 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{5}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

\begin{figure}
\centering
\includegraphics{paper_comparison_files/figure-latex/power-com-1.pdf}
\caption{\label{fig:power-com}power com}
\end{figure}
